Hydra Battles and Cie

### 2021-02-10 doc/chapter-powers.tex
creation of a multiple index

\subsubsection{Using 2x2 integer matrices}
\index{Fibonacci numbers!Matrix exponentiation}

@@

\index{Exercises}

@@

\index{Exercises}
\begin{exercise}
  Prove formally these properties. \emph{Please note that their proof is not needed in our development, they just help to understand the following optimization.}
\end{exercise}

@@

\subsection{A common notation for multiplication}
\label{op-classes}
\index{Coq!Type Classes!Operational Type Classes}
\emph{Operational type classes}~\cite{BS2011}
allow us to define a common notation 
for multiplication in any algebraic structure. 
First, we associate a class to the notion of \emph{multiplication} 
on any type $A$.

@@

\subsection{The Monoid Type Class}
\index{Coq!Type Classes}
We are now ready to  give a definition of the \texttt{Monoid} class, using
\texttt{*} as an infix operator in scope \coqscope{M} for the monoid  multiplication.

@@

\subsection{Monoids and Equivalence Relations}
\index{Coq!Generalized rewriting}
\index{Coq!Equivalence relations}

@@

\index{Coq!Type Classes}
\index{Coq!Proper}
\label{EMonoid-def}

@@

\index{Coq!Coercions}

@@

In order to make possible proof by rewriting on expressions which contain
the exponentiation operator, we have to prove that, whenever \texttt{$x$ == $y$},
the equality \texttt{$x^n$ == $y^n$} holds for any exponent \texttt{$n$}. 
For this purpose, we use the \texttt{Proper} class of module
\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Classes.Morphisms.html}{Coq.Classes.Morphisms}
\index{Coq!Type Classes}
\index{Coq!Proper}
\begin{Coqsrc}
Global Instance power_proper :
     Proper (equiv ==> eq ==> equiv) power.
(* Proof omitted *)
\end{Coqsrc}

@@

\begin{itemize}
\item Continuation Passing Style (CPS) \index{Programming Styles!Continuation Passing Style (CPS)} \cite{reynolds93}
is a way to make explicit the 
     control in the evaluation of an expression, in a purely functional way. 
    For every intermediate computation step, the result is sent
    to a \emph{continuation} that executes the further continuations.
   When the continuation is a lambda-abstraction, its bound variable 
   gives a \emph{name} to this result

\item Like in Parametric Higher Order Abstract Syntax (PHOAS)~\cite{PHOAS}, \index{Parametric Higher-Order Abstract Syntax (PHOAS)}
     the local variables associated to intermediate results are
     represented by variables of  type $A$, where $A$ is the underlying type
  of the considered monoid.
\end{itemize}

@@

\index{Projects}
\begin{project}
Study how  to compile efficiently such data structures.

@@

\index{Projects}
\begin{project}
Define a function which returns the sequence of operations defined by a chain.
For instance, the chain \texttt{C87} of Figure \ref{fig:C87} can be represented as a 
list containing terms of the form \texttt{($i$, Add $j$ $k$)} whenever the associated computation contains the operation $x^i=x^j\times x^k$.

@@

\subsection{Correctness Proofs by Reflection}
\label{reflection-section}
\index{Coq!Proof by reflection}
Instead of letting the tactic \texttt{rewrite} look for contexts in which
setoid rewriting is possible, we propose to use (deterministic) computations for
obtaining a ``canonical'' form for terms generated from a variable \texttt{x}
by contructors associated with monoid multiplication and neutral element.

@@

\subsubsection{About parametricity}
\index{Coq!Parametricity}
Let us now present another tactic for proving chain correctness,
in the tradition of works on \emph{parametricity} and its use for 
proving properties on programs.
Strachey~\cite{Strachey:2000:FCP:609150.609208}
explores the nature of \emph{parametric
polymorphism}: ``\emph{Polymorphic functions behave uniformly for all types}''
then Reynolds~\cite{REYNOLDS83} formalizes this notion through binary relations.
Wadler~\cite{Wadler1989}, then Cohen \emph{et al.}~\cite{Cohen2013}
use this relation for deriving
 theorems about functions that operate on parametric
polymorphic types.

@@

@@

\index{Exercises}
\begin{exercise}
Prove that for any positive integer $p$,  the length of any optimal chain 
for $p$ is less  than twice the number of digits of the binary representation of $p$.
\end{exercise}

@@

\subsubsection{Definition:} 
\index{Coq!Type Classes}
\index{Coq!Proper}
A continuation \texttt{k:Fkont A} is \emph{proper}
if, whenever \linebreak[3] \texttt{x == y} holds, the computations \texttt{k x} and 
\texttt{k y} are equivalent.

@@

\index{Programming Styles!Continuation Passing Style (CPS)}

@@

The following dependently-typed functions will help us to specify  formally
any correct chain generator.
\index{Coq!Dependently Typed Functions}

@@

\index{Projects}
\begin{project}[Optimality and relative efficiency]

>>>>

\subsubsection{Using 2x2 integer matrices}
\index{maths}{Fibonacci numbers!Matrix exponentiation}

@@

\index{additions}{Exercises}

@@

\index{additions}{Exercises}
\begin{exercise}
  Prove formally these properties. \emph{Please note that their proof is not needed in our development, they just help to understand the following optimization.}
\end{exercise}

@@

\subsection{A common notation for multiplication}
\label{op-classes}
\index{coq}{Type classes!Operational type classes}
≈Ò
\emph{Operational type classes}~\cite{BS2011}
allow us to define a common notation 
for multiplication in any algebraic structure. 
First, we associate a class to the notion of \emph{multiplication} 
on any type $A$.

@@

\subsection{The Monoid Type Class}
\index{coq}{Type classes}
We are now ready to  give a definition of the \texttt{Monoid} class, using
\texttt{*} as an infix operator in scope \coqscope{M} for the monoid  multiplication.

@@

\subsection{Monoids and Equivalence Relations}
\index{coq}{Generalized rewriting}
\index{coq}{Type classes!Equivalence relations}

@@

\index{coq}{Type classes}
\index{Coq}{Type classes!Proper class}
\label{EMonoid-def}

@@

\index{coq}{Coercions}

@@

In order to make possible proof by rewriting on expressions which contain
the exponentiation operator, we have to prove that, whenever \texttt{$x$ == $y$},
the equality \texttt{$x^n$ == $y^n$} holds for any exponent \texttt{$n$}. 
For this purpose, we use the \texttt{Proper} class of module
\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Classes.Morphisms.html}{Coq.Classes.Morphisms}
\index{coq}{Type classes}
\index{coq}{Type classes!Proper class}
\begin{Coqsrc}
Global Instance power_proper :
     Proper (equiv ==> eq ==> equiv) power.
(* Proof omitted *)
\end{Coqsrc}

@@

\begin{itemize}
\item Continuation Passing Style (CPS) 
\index{coq}{Continuation Passing Style (CPS)} \cite{reynolds93}
is a way to make explicit the 
     control in the evaluation of an expression, in a purely functional way. 
    For every intermediate computation step, the result is sent
    to a \emph{continuation} that executes the further continuations.
   When the continuation is a lambda-abstraction, its bound variable 
   gives a \emph{name} to this result

\item Like in Parametric Higher Order Abstract Syntax (PHOAS)~\cite{PHOAS}, \index{coq}{Parametric Higher-Order Abstract Syntax (PHOAS)}
     the local variables associated to intermediate results are
     represented by variables of  type $A$, where $A$ is the underlying type
  of the considered monoid.
\end{itemize}

@@

\index{additions}{Projects}
\begin{project}
Study how  to compile efficiently such data structures.

@@

\index{additions}{Projects}
\begin{project}
Define a function which returns the sequence of operations defined by a chain.
For instance, the chain \texttt{C87} of Figure \ref{fig:C87} can be represented as a 
list containing terms of the form \texttt{($i$, Add $j$ $k$)} whenever the associated computation contains the operation $x^i=x^j\times x^k$.

@@

\subsection{Correctness Proofs by Reflection}
\label{reflection-section}
\index{coq}{Proof by reflection}
Instead of letting the tactic \texttt{rewrite} look for contexts in which
setoid rewriting is possible, we propose to use (deterministic) computations for
obtaining a ``canonical'' form for terms generated from a variable \texttt{x}
by contructors associated with monoid multiplication and neutral element.

@@

\subsubsection{About parametricity}
\index{coq}{Parametricity}
Let us now present another tactic for proving chain correctness,
in the tradition of works on \emph{parametricity} and its use for 
proving properties on programs.
Strachey~\cite{Strachey:2000:FCP:609150.609208}
explores the nature of \emph{parametric
polymorphism}: ``\emph{Polymorphic functions behave uniformly for all types}''
then Reynolds~\cite{REYNOLDS83} formalizes this notion through binary relations.
Wadler~\cite{Wadler1989}, then Cohen \emph{et al.}~\cite{Cohen2013}
use this relation for deriving
 theorems about functions that operate on parametric
polymorphic types.

@@

\index{coq}{Plug-ins!paramcoq}

@@

\index{additions}{Exercises}
\begin{exercise}
Prove that for any positive integer $p$,  the length of any optimal chain 
for $p$ is less  than twice the number of digits of the binary representation of $p$.
\end{exercise}

@@

\subsubsection{Definition:} 
\index{coq}{Type classes}
\index{coq}{Type classes!Proper class}
A continuation \texttt{k:Fkont A} is \emph{proper}
if, whenever \linebreak[3] \texttt{x == y} holds, the computations \texttt{k x} and 
\texttt{k y} are equivalent.

@@

\index{coq}{Continuation Passing Style (CPS)}

@@

The following dependently-typed functions will help us to specify  formally
any correct chain generator.
\index{coq}{Dependently typed functions}

@@

\index{additions}{Projects}
\begin{project}[Optimality and relative efficiency]

### 2021-02-10 doc/part-hydras.tex
add a new exercise
fix errors
creation of a multiple index

\section*{Future Work (projects)}
\index{Projects}

@@

%\index{To do}
\index{Projects}

@@

\index{Projects}

@@

\index{Exercises}

@@

\index{Coq!Mutually inductive types}
\index{Coq!Commands!Scheme}

@@

\index{Predicates!round\_n}

@@

\index{Predicates!round}
\label{sect:infix-round}
\begin{Coqsrc}
Definition round h h' := exists n,  round_n n h h'.

@@

\index{Predicates!Termination}

@@

\subsection{A  Small Proof of Impossibility}
\index{Maths!Proofs of impossibility}

@@

The next chapter is dedicated to a generic formalization of ordinal notations, and chapter~\ref{chap:T1} to a proof of termination of all hydra battles with the help of an ordinal notation for the interval $[0,\epsilon_0)$\,\footnote{We use the mathematical notation $[a,b)$ for the interval $\{x|a\leq x < b\}$.}.
\index{Maths!Notations!Interval}

@@

\index{Maths!Ordinal numbers}

@@

\index{Exercises}
\begin{exercise}
Prove, that, in any ordinal notation system, every ordinal has at most one predecessor, and at most one successor.

@@

\index{Exercises}
\begin{exercise}
Prove, that, in any ordinal notation system, if $\beta$ is a successor of $\alpha$,
then for any $\gamma$, $\gamma<\beta$ implies 
$\gamma\leq\alpha$.

@@

\section{The Ordinal \texorpdfstring{$\omega^2$}{omega^2}}

@@

\index{Coq!Commands!Function}

@@

\index{Maths!Transfinite induction}

@@

\index{Coq!Commands!Program}

@@

\index{Exercises}
\begin{exercise}
Prove that \texttt{Omega\_plus\_Omega} cannot be a sub-notation of \texttt{Omega}.
\end{exercise}

\index{Projects}
\begin{project}
Adapt the definition of \texttt{Hvariant} (Sect.~\ref{sect:hvariant-def}) in order to
have an ordinal notation as argument. Prove that if $O_A$ is a sub-notation of $O_B$, then any variant defined on  $O_A$ can be automatically transformed into 
a variant on $O_B$.
\end{project}

@@

\index{Projects}
\label{exo:ON-mult}
\begin{project}
% Define in \coq{} the product of two ordinal notations $N_A$ and $N_B$.
% If $A$ [resp. $B$] is the underlying type of $N_A$ [resp. $N_B$], the
% product \texttt{ON\_mult $N_A$ $N_B$} is implemented over the cartesian product $B\times A$ (with the lexicographic ordering).

@@

\index{Projects}
\begin{project}
Consider two isomorphic ordinal notations \texttt{OA} and \texttt{OB}.
Prove that, if \texttt{OA} [resp. \texttt{OB}] is a correct implementation 
of $\alpha$ [resp. $\beta$], then $\alpha=\beta$.
\end{project}

\index{Projects}
\begin{project}
\label{project:succ-limit-dec}
Add to the class \texttt{ON} the requirement that for any $\alpha$ it is decidable whether $\alpha$ is $0$, a successor or a limit ordinal.

@@

\index{Projects}
\begin{project}
\label{project:on-setoid}
Reconsider the  class \texttt{ON}, with an equivalence instead of Leibniz equality.
\end{project}

@@

\subsection{Cantor Normal Form}
\index{Maths!Cantor normal form}

@@

\index{Notations!phi0@phi0 (exponential of base omega)}

@@

\index{Maths!Additive principal ordinals}

@@

\label{types:ppT1}
\index{Functions!pp@ pp (pretty printing terms in Cantor normal form)}

@@

\index{Projects}
\begin{project}
Design  (in \ocaml?) a set of tools for systematically pretty printing ordinal terms in Cantor normal form.
\end{project}

@@

\index{Coq!Unicity of equality proofs}

@@

\index{Exercises}
\begin{exercise}
In earlier versions of this development, the predicate \texttt{nf} was defined  inductively, with various constructors describing all possible cases.
\begin{enumerate}
\item Please give such a definition, in a dedicated module.
\item Prove the logical equivalence between your definition and ours.
\item Define a variant of the type \texttt{E0} (with your definition of \texttt{nf}).
\item Can you still prove a lemma like \texttt{E0\_eq\_iff} ? With the help of an axiom from some module of the standard library ?
\end{enumerate}
\end{exercise}

@@

\index{Exercises}
 \begin{exercise}
Prove in \coq{} that for any ordinal $0< \alpha<\epsilon_0$, $\alpha$ is a limit if 
and only if for all $\beta<\alpha$, the interval $[\beta,\alpha)$ is
infinite.

@@

%% To simplify !
\index{Exercises}
\begin{exercise}
Prove that for any ordinal $\alpha:\texttt{E0}$, 
$\omega\leq \alpha$ if and only if, for any natural number $i$,
$i+\alpha=\alpha$.

@@

\subsubsection{A First Attempt}
\label{sec:orgheadline77}
\index{Coq!Well-founded induction}

@@

\index{Projects}
\begin{project}
 \emph{This exercise is a continuation of Project~\vref{exo:ON-mult}.}
Use \texttt{ON\_mult} to define an ordinal notation \texttt{Omega2} for $\omega^2=\omega\times\omega$.

@@

\index{Projects}
\begin{project}
The class \texttt{ON} of ordinal notations has been defined long after this 
chapter, and is not used yet in the development of the type \texttt{E0}. 
A better integration of both notions should simplify the development on ordinals in Cantor normal form. This integration is planned for the future versions.

@@

\index{Projects}
\begin{project}
Compare \texttt{oplus}'s definition (with inner fixpoint) with other possibilities
(\texttt{coq-equations}, \texttt{Function}, etc.).
\end{project}
\subsection{More Theorems on Hessenberg's Sum}

@@

\chapter[The Ketonen-Solovay machinery]{Accessibility inside \texorpdfstring{$\epsilon_0$}{Epsilon0}: The Ketonen-Solovay Machinery\label{ks-chapter}}
\label{chap:ketonen}
\index{Maths!Ketonen-Solovay machinery}

@@

\section{Canonical Sequences}
\label{ketonen-solovay-sect}
\index{Maths!Canonical sequences}

@@

\index{Maths!Transfinite induction}
\begin{itemize}
\item If $\alpha\not=0$, then $\canonseq{\alpha}{i}<\alpha$. Thus canonical sequences can be used for proofs by transfinite induction or function definition by transfinite recursion
\item If $\lambda$ is a limit ordinal, then $\lambda$ is the least upper bound of the set 
$\{\canonseq{\lambda}{i}\;|\,i\in\mathbb{N}_1\}$

@@

\index{Projects}
\begin{project}
Many lemmas presented in this chapter were stated and proved before the introduction of 
the type class \texttt{ON} of ordinal notations, and in particular its  instance \texttt{Epsilon0}.
Thus definitions and lemmas refer to the type \texttt{T1} of possibly not well-formed terms.
This should be fixed in  a future version.
\end{project}

@@

\textbf{Hint:} You may add to your function a third argument usually called \texttt{fuel} for allowing you to give a structurally 
recursive function (\emph{cf} the post of Guillaume Melquiond on Coq-club (Dec 21, 2020)
\url{https://sympa.inria.fr/sympa/arc/coq-club/2020-12/msg00069.html}).
The type \texttt{fuel}, an alternative 
to \texttt{nat} is available on \href{../theories/html/hydras.Prelude.Fuel.html}{Prelude.Fuel}).
\index{Coq!Giving fuel to a long computation}

@@

\section{Accessibility inside \texorpdfstring{$\epsilon_0$}{epsilon0} : Paths}
\index{Maths!Accessibility inside epsilon0}
\label{sect:pathes-intro}

@@

\index{Predicates!path\_to}
\label{sect:path-to-def}

@@

\subsubsection{Sketch of Proof of \texttt{Cor12}}
\index{Maths!Transfinite induction}

@@

\index{Exercises}
\begin{exercise}
  Prove that  the length of the $i+1$-path from
  $\omega^\omega$ to $\omega^i$ is $1 + (i+1)^{(i+1)}$, for any $i$. Note that the $i$-path from
  $\omega^\omega$ to $\omega^i$ is only one step long.
 \end{exercise}

@@

\label{lemma:L-2_6-1}
\index{Coq!Sigma types}

@@

\index{Projects}
\begin{project}
Study a possible modification of the definition of a variant  (for  standard battles).

@@

\index{Maths!Large sequences}
\index{Maths!Minimal large sequences}

@@

\index{Predicates!mlarge @ mlarge (minimal large sequences)}
\begin{Coqsrc}
Definition mlarge alpha (s:list nat) := path_to zero s alpha.
\end{Coqsrc}

@@

\index{Coq!Plug-ins!Equations}

@@

\index{Projects}
\begin{project}
Instead of considering standard paths and battles, consider ``constant'' paths and the corresponding battles. Please use \texttt{Equations} in order to define the function that computes the length of the $k$-path which leads  from $\alpha$ to $0$.
Prove a few  exact formulas and minoration lemmas.
\end{project}

@@

\label{sect:hardy}
In order to give an idea of the complexity of the functions  $L_\alpha$s, we compare them with a better known family of functions, the so called \emph{Wainer-Hardy hierarchy} of fast growing functions,
presented for instance in~\cite{Promel2013}. 
\index{Maths!Hardy Hierarchy}

@@

\index{Coq!Plug-ins!Equations}
\label{Functions:H-alpha}

@@

\index{Maths!Wainer Hierarchy}

@@

\index{Coq!Plug-ins!Equations}
\begin{Coqsrc}
Fail Equations F_ (alpha: E0) (i:nat) :  nat  by wf  alpha Lt :=
  F_ alpha  i with E0_eq_dec alpha Zero :=
    { | left _ =>  i ;
      | right nonzero
          with Utils.dec (Limitb alpha) :=
          { | left _ =>  F_ (Canon alpha i)  i ;
            | right notlimit =>  iterate (F_ (Pred alpha))  (S i) i}}.
\end{Coqsrc}

@@

\index{Coq!Plug-ins!Equations}
\label{Functions:F-alpha}
\index{Maths! Fast growing functions}

@@

\index{Maths!Ordering functions}

@@

\index{Maths!Additive principal ordinals}

@@

\index{Predicates!Closed}

@@

\index{Maths!Critical ordinals}

@@

Is it true ?
\end{exercise}

@@

\index{Projects}
\begin{project}
There is no construction of ordinal multiplication in~\cite{schutte}. 
It would be interesting to derive this operation from Schütte's axioms,
and prove its consistence with multiplication in ordinal notations for 
$\epsilon_0$ and $\Gamma_0$.
\end{project}

@@

\index{Projects}
\begin{project}
Prove that a (countable)  ordinal $\alpha$ is strongly critical iff 
$\phi_\alpha(0)=\alpha$ (Theorem 13.13 of~\cite{schutte} ). 
\end{project}

\index{Projects}
\begin{project}
Prove that the set of strongly critical ordinals is unbounded and closed (Theorem 13.14 of~\cite{schutte} ). Thus this set is not empty,  hence has a least element. Otherwise, the definition of $\Gamma_0$ above would be useless.
\end{project}

@@

\index{Projects}
\begin{project}
This theorem is not formally proved in this development yet. It should be!
\end{project}

@@

\index{Projects}
\begin{project}
Write a tactic that solves automatically goals of the form (\texttt{$\alpha$ t2< $\beta$}), where $\alpha$ and $\beta$ are closed terms of type \texttt{T2}.
\end{project}

@@

\index{Projects}
\begin{project}
Please study a way to pretty print ordinal terms in Veblen normal form (see Section~\vref{sect:ppT1}).
\end{project}

@@

\index{Projects}
\begin{project}
Prove that \texttt{Epsilon0} (page~\pageref{instance-epsilon0})
is a sub-notation system of \texttt{Gamma0}.

>>>>

\section*{Future Work (projects)}
\index{hydras}{Projects}

@@

%\index{To do}
\index{hydras}{Projects}

@@

\index{hydras}{Projects}

@@

\index{hydras}{Exercises}

@@

\index{coq}{Mutually inductive types}
\index{coq}{Commands!Scheme}

@@

\index{hydras}{Predicates!round\_n}

@@

\index{hydras}{Predicates!round}
\label{sect:infix-round}
\begin{Coqsrc}
Definition round h h' := exists n,  round_n n h h'.

@@

\index{hydras}{Predicates!Termination}

@@

\subsection{A  Small Proof of Impossibility}
\index{coq}{Proofs of impossibility}

@@

The next chapter is dedicated to a generic formalization of ordinal notations, and chapter~\ref{chap:T1} to a proof of termination of all hydra battles with the help of an ordinal notation for the interval $[0,\epsilon_0)$\,\footnote{We use the mathematical notation $[a,b)$ for the interval $\{x|a\leq x < b\}$.}.
\index{maths}{Notations!Interval}

@@

\index{maths}{Ordinal numbers}

@@

\index{hydras}{Exercises}
\begin{exercise}
Prove, that, in any ordinal notation system, every ordinal has at most one predecessor, and at most one successor.

@@

\index{hydras}{Exercises}
\begin{exercise}
Prove, that, in any ordinal notation system, if $\beta$ is a successor of $\alpha$,
then for any $\gamma$, $\gamma<\beta$ implies 
$\gamma\leq\alpha$.

@@

\section{The Ordinal \texorpdfstring{$\omega^2$}{omega2}}

@@

\index{coq}{Commands!Function}

@@

\index{maths}{Transfinite induction}

@@

\index{coq}{Commands!Program}

@@

\index{hydras}{Exercises}
\begin{exercise}
Prove that \texttt{Omega\_plus\_Omega} cannot be a sub-notation of \texttt{Omega}.
\end{exercise}

\index{hydras}{Projects}
\begin{project}
Adapt the definition of \texttt{Hvariant} (Sect.~\ref{sect:hvariant-def}) in order to
have an ordinal notation as argument. Prove that if $O_A$ is a sub-notation of $O_B$, then any variant defined on  $O_A$ can be automatically transformed into 
a variant on $O_B$.
\end{project}

@@

\index{hydras}{Projects}
\label{exo:ON-mult}
\begin{project}
% Define in \coq{} the product of two ordinal notations $N_A$ and $N_B$.
% If $A$ [resp. $B$] is the underlying type of $N_A$ [resp. $N_B$], the
% product \texttt{ON\_mult $N_A$ $N_B$} is implemented over the cartesian product $B\times A$ (with the lexicographic ordering).

@@

\index{hydras}{Projects}
\begin{project}
Consider two isomorphic ordinal notations \texttt{OA} and \texttt{OB}.
Prove that, if \texttt{OA} [resp. \texttt{OB}] is a correct implementation 
of $\alpha$ [resp. $\beta$], then $\alpha=\beta$.
\end{project}

\index{hydras}{Projects}
\begin{project}
\label{project:succ-limit-dec}
Add to the class \texttt{ON} the requirement that for any $\alpha$ it is decidable whether $\alpha$ is $0$, a successor or a limit ordinal.

@@

\index{hydras}{Projects}
\begin{project}
\label{project:on-setoid}
Reconsider the  class \texttt{ON}, with an equivalence instead of Leibniz equality.
\end{project}

@@

\subsection{Cantor Normal Form}
\index{maths}{Cantor normal form}

@@

\index{hydras}{Notations!phi0@phi0 (exponential of base omega)}

@@

\index{maths}{Additive principal ordinals}

@@

\label{types:ppT1}
\index{hydras}{Functions!pp@ pp (pretty printing terms in Cantor normal form)}

@@

\index{hydras}{Projects}
\begin{project}
Design  (in \ocaml?) a set of tools for systematically pretty printing ordinal terms in Cantor normal form.
\end{project}

@@

\index{coq}{Unicity of equality proofs}

@@

\index{hydras}{Exercises}
\begin{exercise}
In earlier versions of this development, the predicate \texttt{nf} was defined  inductively, with various constructors describing all possible cases.
\begin{enumerate}
\item Please give such a definition, in a dedicated module.
\item Prove the logical equivalence between your definition and ours.
\item Define a variant of the type \texttt{E0} (with your definition of \texttt{nf}).
\item Can you still prove a lemma like \texttt{E0\_eq\_iff} ? With the help of an axiom from some module of the standard library ?
\end{enumerate}
\end{exercise}

@@

\index{hydras}{Exercises}
 \begin{exercise}
Prove in \coq{} that for any ordinal $0< \alpha<\epsilon_0$, $\alpha$ is a limit if 
and only if for all $\beta<\alpha$, the interval $[\beta,\alpha)$ is
infinite.

@@

%% To simplify !
\index{hydras}{Exercises} 
\begin{exercise}
Prove that for any ordinal $\alpha:\texttt{E0}$, 
$\omega\leq \alpha$ if and only if, for any natural number $i$,
$i+\alpha=\alpha$.

@@

\subsubsection{A First Attempt}
\label{sec:orgheadline77}
\index{coq}{Well-founded induction}

@@

\index{hydras}{Projects}
\begin{project}
 \emph{This exercise is a continuation of Project~\vref{exo:ON-mult}.}
Use \texttt{ON\_mult} to define an ordinal notation \texttt{Omega2} for $\omega^2=\omega\times\omega$.

@@

\index{hydras}{Projects}
\begin{project}
The class \texttt{ON} of ordinal notations has been defined long after this 
chapter, and is not used yet in the development of the type \texttt{E0}. 
A better integration of both notions should simplify the development on ordinals in Cantor normal form. This integration is planned for the future versions.

@@

\index{hydras}{Projects}
\begin{project}
Compare \texttt{oplus}'s definition (with inner fixpoint) with other possibilities
(\texttt{coq-equations}, \texttt{Function}, etc.).
\end{project}
\subsection{More Theorems on Hessenberg's Sum}

@@

\chapter[The Ketonen-Solovay machinery]{Accessibility inside \texorpdfstring{$\epsilon_0$}{Epsilon0}: The Ketonen-Solovay Machinery\label{ks-chapter}}
\label{chap:ketonen}
\index{maths}{Ordinal numbers!Ketonen-Solovay machinery}

@@

\section{Canonical Sequences}
\label{ketonen-solovay-sect}
\index{maths}{Ordinal numbers!Canonical sequences}

@@

\index{maths}{Transfinite induction}
\begin{itemize}
\item If $\alpha\not=0$, then $\canonseq{\alpha}{i}<\alpha$. Thus canonical sequences can be used for proofs by transfinite induction or function definition by transfinite recursion
\item If $\lambda$ is a limit ordinal, then $\lambda$ is the least upper bound of the set 
$\{\canonseq{\lambda}{i}\;|\,i\in\mathbb{N}_1\}$

@@

\index{hydras}{Projects}
\begin{project}
Many lemmas presented in this chapter were stated and proved before the introduction of 
the type class \texttt{ON} of ordinal notations, and in particular its  instance \texttt{Epsilon0}.
Thus definitions and lemmas refer to the type \texttt{T1} of possibly not well-formed terms.
This should be fixed in  a future version.
\end{project}

@@

\textbf{Hint:} You may add to your function a third argument usually called \texttt{fuel} for allowing you to give a structurally 
recursive function (\emph{cf} the post of Guillaume Melquiond on Coq-club (Dec 21, 2020)
\url{https://sympa.inria.fr/sympa/arc/coq-club/2020-12/msg00069.html}).
The type \texttt{fuel}, an alternative 
to \texttt{nat} is available on \href{../theories/html/hydras.Prelude.Fuel.html}{Prelude.Fuel}).
\index{coq}{Giving fuel to a long computation}

@@

\section{Accessibility inside \texorpdfstring{$\epsilon_0$}{epsilon0} : Paths}
\index{maths}{Ordinal numbers!Accessibility inside epsilon0}
\label{sect:pathes-intro}

@@

\index{hydras}{Predicates!path\_to}
\label{sect:path-to-def}

@@

\subsubsection{Sketch of Proof of \texttt{Cor12}}
\index{maths}{Transfinite induction}

@@

\index{hydras}{Exercises}
\begin{exercise}
  Prove that  the length of the $i+1$-path from
  $\omega^\omega$ to $\omega^i$ is $1 + (i+1)^{(i+1)}$, for any $i$. Note that the $i$-path from
  $\omega^\omega$ to $\omega^i$ is only one step long.
 \end{exercise}

@@

\label{lemma:L-2_6-1}
\index{coq}{Sigma types}

@@

\index{hydras}{Projects}
\begin{project}
Study a possible modification of the definition of a variant  (for  standard battles).

@@

\index{maths}{Ordinal numbers!Large sets}
\index{maths}{Ordinal numbers!Minimal large sets}

@@

\index{hydras}{Predicates!mlarge @ mlarge (minimal large sequences)}
\begin{Coqsrc}
Definition mlarge alpha (s:list nat) := path_to zero s alpha.
\end{Coqsrc}

@@

\index{coq}{Plug-ins!Equations}

@@

\index{hydras}{Projects}
\begin{project}
Instead of considering standard paths and battles, consider ``constant'' paths and the corresponding battles. Please use \texttt{Equations} in order to define the function that computes the length of the $k$-path which leads  from $\alpha$ to $0$.
Prove a few  exact formulas and minoration lemmas.
\end{project}

@@

\label{sect:hardy}
In order to give an idea of the complexity of the functions  $L_\alpha$s, we compare them with a better known family of functions, the so called \emph{Wainer-Hardy hierarchy} of fast growing functions,
presented for instance in~\cite{Promel2013}. 
\index{maths}{Rapidly growing functions!Hardy Hierarchy}

@@

\index{coq}{Plug-ins!Equations}
\label{Functions:H-alpha}

@@

\index{maths}{Rapidly growing functions!Wainer Hierarchy}

@@

\index{coq}{Plug-ins!Equations}
\begin{Coqsrc}
Fail Equations F_ (alpha: E0) (i:nat) :  nat  by wf  alpha Lt :=
  F_ alpha  i with E0_eq_dec alpha Zero :=
    { | left _ =>  i ;
      | right nonzero
          with Utils.dec (Limitb alpha) :=
          { | left _ =>  F_ (Canon alpha i)  i ;
            | right notlimit =>  iterate (F_ (Pred alpha))  (S i) i}}.
\end{Coqsrc}

@@

\index{coq}{Plug-ins!Equations}
\label{Functions:F-alpha}
\index{maths}{Rapidly growing functions}

@@

\index{maths}{Ordinal numbers!Ordering functions}

@@

\index{maths}{Ordinal numbers!Additive principal ordinals}

@@

\index{hydras}{Predicates!Closed}

@@

\index{maths}{Ordinal numbers!Critical ordinals}

@@

Is it true ?

\emph{You may start this exercise with the file
\url{../exercises/ordinals/schutte_cnf_counter_example.v}.}
\end{exercise}

@@

\index{hydras}{Projects}
\begin{project}
There is no construction of ordinal multiplication in~\cite{schutte}. 
It would be interesting to derive this operation from Schütte's axioms,
and prove its consistence with multiplication in ordinal notations for 
$\epsilon_0$ and $\Gamma_0$.
\end{project}

@@

\index{hydras}{Projects}
\begin{project}
Prove that a (countable)  ordinal $\alpha$ is strongly critical iff 
$\phi_\alpha(0)=\alpha$ (Theorem 13.13 of~\cite{schutte} ). 
\end{project}

\index{hydras}{Projects}
\begin{project}
Prove that the set of strongly critical ordinals is unbounded and closed (Theorem 13.14 of~\cite{schutte} ). Thus this set is not empty,  hence has a least element. Otherwise, the definition of $\Gamma_0$ above would be useless.
\end{project}

@@

\index{hydras}{Projects}
\begin{project}
This theorem is not formally proved in this development yet. It should be!
\end{project}

@@

\index{hydras}{Projects}
\begin{project}
Write a tactic that solves automatically goals of the form (\texttt{$\alpha$ t2< $\beta$}), where $\alpha$ and $\beta$ are closed terms of type \texttt{T2}.
\end{project}

@@

\index{hydras}{Projects}
\begin{project}
Please study a way to pretty print ordinal terms in Veblen normal form (see Section~\vref{sect:ppT1}).
\end{project}

@@

\index{hydras}{Projects}
\begin{project}
Prove that \texttt{Epsilon0} (page~\pageref{instance-epsilon0})
is a sub-notation system of \texttt{Gamma0}.
